---
title: Using tidyr::pivot_longer() and regex for data wrangling
author: ~
date: '2020-03-16'
slug: using-pivot-longer-and-regex-for-data-wrangling
type: blog
categories: []
tags: []
Categories:
  - Data wrangling
  - regex
Description: ''
Tags:
  - Data wrangling
  - regex
menu: []
---

# TL;DR

This code shows you how you can effectively wrangle your dataframe from wide to 
long using `tidyr::pivot_longer()` combined with regular expressions for properly 
naming the variables. 

# Your data is... not in the right format!

You have a wide dataset of patient data, each patient identified through `ID`. 
They have been hospitalised multiple times, each place of hospitalisation and 
outcome listed in its own variable. Place and Outcome of a hospitalization are 
linked with the same suffix number, e.g. `_2_2`: 

```{r}
# load packages
suppressPackageStartupMessages(library(tidyverse))

# create sample data
patient_data <- 
tibble::tribble(~ID, ~Hosp_Place_1, ~Hosp_Outcome_1, ~Hosp_Place_2_2, 
                ~Hosp_Outcome_2_2, ~Hosp_Place_2_3, ~Hosp_Outcome_2_3, 
                1, "London", "Alive", "Paris", "alive", "Rome", "dead",
                2, "Paris", "alive", "Rome", "alive", "London", "alive",
                3, "Berne", "dead", NA_character_, NA_character_, NA_character_, NA_character_) 

# look at data
patient_data

```

Patient No. 3 died in his first hospitatisation, so the following variables are listed `NA`.  

For some analysis, this data structure would work ok, but for your specific questions 
you need to switch the unit of analysis to "hospitalisation" instead of "patient". 
In other words, you'd like to switch the data from wide to long. 

Thanks to the tidyverse's initiative of making clear function names, it might be 
obvious, that `tidyr::pivot_longer()` should do what you want.  

# Familiarize yourself with the function - What to do with tidyr::pivot_longer()?

The first step is to initially think about, what the outcome should look like and 
what input-arguments the function takes, so let's do this: 

```{r, eval=FALSE}
# tidyr::pivot_long() and its arguments
tidyr::pivot_longer(data, 
                    cols, 
                    names_to = "name", 
                    names_prefix = NULL, 
                    names_sep = NULL, 
                    names_pattern = NULL,
                    names_ptypes = list(),
                    names_repair = "check_unique",
                    values_to = "value",
                    values_drop_na = FALSE,
                    values_ptypes = list())
```

As you can see, only `data` and `cols` are effectively needed, so let's try that. 
As I want to exclude the Patient-ID from pivoting, I remove this line from pivoting: 

```{r}
# as with all the tidyverse functions you can easily pipe-in the data as 
# the first argument

# '-ID' means, that all variables are used other than ID
patient_data %>% 
  pivot_longer(cols = -ID)

# I could have put the following instead, meaning selecting the variables I want
# but it was generally shorter to drop just the ID
# patient_data %>%
#   pivot_longer(cols = Hosp_Place_1:Hosp_Outcome_2_3)

```

Something clearly happened, but `name` and `value` are not exactly what we want here.  
The colnames "name" and "value" are actually coming from the default arguments.  
What is now the next step?  

# Mental image of desired outcome - How should my dataframe look like?

My desired output is a dataframe with the colum names `ID`, `Hosp_Place` and `Hosp_Outcome`. 
Additionally, I want a variable - lets call it `hosp_sequence` - that captures the number of 
the hospitalisation (you remember the suffix of the original variable names).  

`tidyr::pivot_longer()`'s `names_to =`-arguments states in the help-page:  
*Can be a character vector, creating multiple columns, if names_sep or names_pattern is provided.*    

If you can detect any patterns in the column names, its possible to use them for the 
column naming.  If we look at Hosp_Place_1 and Hosp_Outcome_1 we can clearly see 
a pattern: The information I want as name is `Hosp_Place` and `Hosp_Outcome` and the 
number followed should be put in variable called `hosp_sequence`.  

This translates to something like `(Hosp_Place)_(1)` where the parts in brackets 
correspond to the inputs given in `names_to = `. With the `.value`-argument, I 
can easily take over the string as it is.  

I actually found this very confusing (honestly - still do...) and I had great help for defining the 
regular expression from R4DS^[https://rfordatascience.slack.com/archives/C8K09CDNZ/p1584129595187200 If you are not already on this slack - sign up for it! Its just so great, low key 
help and great learning opportunities to just dive through the topics. ].  

Finally, this is the code we need:  

```{r}
# this code should do the trick
patient_data %>% 
  pivot_longer(cols = -ID, 
               names_to = c(".value", "hosp_sequence"),
               names_pattern = '(^[A-z]+_[A-z]+)_([0-9].*)')
```


Wait, what is happening with rows 8 and 9? This is my deceased Patient No. 3 an those 
are not hospitalisations anymore - how do I drop those rows? 

```{r}
# you can either use dplyr::drop_na() or specify the built-in argument to TRUE
patient_data %>% 
  pivot_longer(cols = -ID, 
               names_to = c(".value", "hosp_sequence"),
               names_pattern = '(^[A-z]+_[A-z]+)_([0-9].*)', 
               values_drop_na = TRUE)
```

# Comment

I wrote this blogpost after after solving exactly this issue with a real dataset for a colleague.  
My work as a research programmer allows me to dive into data wrangling problems on a regular basis. 
As I learned most of my R skills from other blogposts from the fantastic R community, I 
started to write up some of the problems I encountered for others.  I also use my 
previous blogposts sometimes, when I have to dig up old code... 

Any comments from your side? Let me know!

